Sequential Composition for Relaxed Memory: Pomsets with Predicate Transformers

### 2021-07-06 paper/intro.tex
got the definitions to line up with using weakest preconditions.

In certain cases it is possible to ban concurrent access using separation
\cite{OHearn:2007:RCL:1235896.1236121}, or to accept inefficient
implementation in order to obtain sequential consistency
\cite{DBLP:conf/snapl/MarinoMMNS15}.  When these approaches are not
available, however, we are left with an enormous gap in our understanding of
one of the most basic elements of computing: the humble semicolon.  Until
recently, existing approaches either
\begin{itemize}
\item didn't bother tracking dependencies, allowing ``thin air'' executions
  \cite{Batty:2011:MCC:1926385.1926394},
\item tracked dependencies conservatively, using syntax, requiring inefficient
  implementation of relaxed access
  \cite{DBLP:conf/pldi/LahavVKHD17,Boehm:2014:OGA:2618128.2618134,DBLP:conf/oopsla/VafeiadisN13,DBLP:journals/corr/abs-1804-04214}---while
  this may be fine for C, it's a non-started for safe languages like Java,
  which lack C's ``plain'' access,
\item computed dependencies using non-compositional operational models over
  alternate worlds \cite{Manson:2005:JMM:1047659.1040336,
    DBLP:conf/esop/JagadeesanPR10,
    DBLP:conf/popl/KangHLVD17,DBLP:journals/pacmpl/ChakrabortyV19,DBLP:conf/pldi/LeeCPCHLV20,promising-ldrf}---as
  discussed in \textsection\ref{sec:promising}, these models fail to validate
  temporal safety properties.
  %JMM, promising semantics, weakestMO, etc
\end{itemize}

@@

In this paper, we show that \PwP{} can be extended with \emph{families of
  predicate transformers} to calculate sequential dependencies in a way that
is \emph{compositional} and \emph{direct}: \emph{compositional} in that the
denotation of $(\aCmd_1\SEMI \aCmd_2)$ can be computed from the denotation of
$\aCmd_1$ and the denotation of $\aCmd_2$, and \emph{direct} in that these
can be calculated independently.

One of the main difficulties is to find a definition that is associative:
the denotation of $((\aCmd_1\SEMI \aCmd_2)\SEMI \aCmd_3)$ should be the same as 
the denotation of $(\aCmd_1\SEMI (\aCmd_2\SEMI \aCmd_3))$.

We have layered the definitions, both to manage complexity and to make clear
that the main contribution of this paper is broadly applicable.  After an
overview, we define sequential dependencies in \textsection\ref{sec:model}.
The next two sections add concurrency.  In \textsection\ref{sec:mca}, we
define \PwTmca{}, which models Java for \mca{} hardware, similar to that of
\citet{DBLP:journals/pacmpl/JagadeesanJR20}.  In \textsection\ref{sec:c11},
we define \PwTc{}, which models \cXI, adapting the approach of
\citet{DBLP:conf/esop/PaviottiCPWOB20}.

We describe two variants of \PwTmca{}.  The first has a nicer semantics; the
second has a more efficient lowering for acquiring reads on \armeight{}.
Both variants have optimal lowering for relaxed access.

>>>>

In certain cases it is possible to ban concurrent access using separation
\cite{OHearn:2007:RCL:1235896.1236121}, or to accept inefficient
implementation in order to obtain sequential consistency
\cite{DBLP:conf/snapl/MarinoMMNS15}.  When these approaches are not
available, however, we are left with an enormous gap in our understanding of
one of the most basic elements of computing: the humble semicolon.  Until
recently, existing approaches either
\begin{itemize}
\item didn't bother tracking dependencies, allowing ``thin air'' executions
  \cite{Batty:2011:MCC:1926385.1926394},
\item tracked dependencies conservatively, using syntax, requiring inefficient
  implementation of relaxed access
  \cite{DBLP:conf/pldi/LahavVKHD17,Boehm:2014:OGA:2618128.2618134,DBLP:conf/oopsla/VafeiadisN13,DBLP:journals/corr/abs-1804-04214}---while
  this may be fine for C, it's a non-started for safe languages like Java,
  which lack C's ``plain'' access,
\item computed dependencies using non-compositional operational models over
  alternate worlds \cite{Manson:2005:JMM:1047659.1040336,
    DBLP:conf/esop/JagadeesanPR10,
    DBLP:conf/popl/KangHLVD17,DBLP:journals/pacmpl/ChakrabortyV19,DBLP:conf/pldi/LeeCPCHLV20,promising-ldrf}---these
  models validate many compiler optimizations, but also
  %allow some dubious executions 
  fail to validate temporal safety properties
  (see \textsection\ref{sec:promising}).
\end{itemize}

@@

In this paper, we show that \PwP{} can be extended with \emph{families of
  predicate transformers (\PwT{})} to calculate sequential dependencies in a way that
is \emph{compositional} and \emph{direct}: \emph{compositional} in that the
denotation of $(\aCmd_1\SEMI \aCmd_2)$ can be computed from the denotation of
$\aCmd_1$ and the denotation of $\aCmd_2$, and \emph{direct} in that these
can be calculated independently.  The definition is associative: the
denotation of $((\aCmd_1\SEMI \aCmd_2)\SEMI \aCmd_3)$ is the same as the
denotation of $(\aCmd_1\SEMI (\aCmd_2\SEMI \aCmd_3))$.  It also validates
expected laws concerning the interaction of sequencing and conditional
execution.

To manage complexity, we have layered the definitions.  After an overview, we
define sequential dependencies in \textsection\ref{sec:model}.  The next two
sections add concurrency.  In \textsection\ref{sec:mca}, we define \PwTmca{},
which models Java for \mca{} hardware, similar to that of
\citet{DBLP:journals/pacmpl/JagadeesanJR20}.  In \textsection\ref{sec:c11},
we define \PwTc{}, which models \cXI, adapting the approach of
\citet{DBLP:conf/esop/PaviottiCPWOB20}.  In \textsection\ref{sec:additional},
we extend the semantics to include additional features, such as address calculation.
In \textsection{sec:results}, we summarize the results.

### 2021-07-06 paper/mca1.tex
monday
got the definitions to line up with using weakest preconditions.

\section{\PwTmcaTITLE{1}: Pomsets with Predicate Transformers for MCA (Part 1)}
\label{sec:mca1}
We define two models for \mca{} architectures: \PwTmca{1} and \PwTmca{2}.
We drop the subscript when referring to both models.

>>>>

\section{\PwTmcaTITLE{}: Pomsets with Predicate Transformers for MCA}
\label{sec:mca}
We define two models for \mca{} architectures: \PwTmca{1} and \PwTmca{2}.
The first has a nicer semantics; the second has a more efficient lowering for
acquiring reads on \armeight{}.  Both variants have optimal lowering for
relaxed access.  We drop the subscript when referring to both models.

\subsection{\PwTmcaTITLE{1}}
\label{sec:mca1}
